### 2022-08-10 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@1924f65396fc26278b64f69f504949fdb8302210 🚀

Understanding the ecological constraints that shape species interactions
networks and predicting their emerging structure are thus two
complementary aims of network ecology. This distinction between
understanding and predicting is essential when using statistical and
mathematical models in network ecology and interpreting them. On one
hand, null models help us identify potential ecological mechanisms that
drive species interactions and constrain ecological networks. Null
models generate a distribution for a target measure using a set of rules
that exclude the mechanism of interest (Fortuna and Bascompte 2006;
Delmas et al. 2019). The difference between the model and empirical data
helps us evaluate the effect of this ecological process in nature. On
the other hand, predictive models can help fill many gaps on species
interactions data. A variety of such models have recently been developed
using machine learning and other statistical tools, most of which are
presented in Strydom et al. (2021). Given the constrained nature of
ecological networks, it is surprising that the principle of maximum
entropy, a rigorous mathematical method designed for both the analysis
and prediction of constrained systems, has been barely used in network
ecology.

@@

The principle of maximum entropy has been used in a wide range of
disciplines, from thermodynamics, chemistry and biology (Martyushev and
Seleznev 2006) to graph and network theory (e.g., Park and Newman 2004;
van der Hoorn, Lippner, and Krioukov 2018). It has also been proven
useful in ecology, e.g.~in species distribution models (Phillips,
Anderson, and Schapire 2006) and macroecological models (Harte et al.
2008; Harte and Newman 2014). In network ecology, MaxEnt has been used
to predict the degree distribution of bipartite networks from the number
of species and the number of interactions (Williams 2011) and to predict
interaction strengths between species pairs using their relative
abundances within an optimal transportation theory regularized with
information entropy (Stock, Poisot, and De Baets 2021). Thus, to the
best of our knowledge, MaxEnt has never been used to predict food-web
structure directly, even though food webs are among the most documented
and widespread ecological networks (Ings et al. 2009).

Food-web properties that can be derived using MaxEnt are varied and
pertain to different elements of the network (i.e.~at the species
(node), the interaction (edge) or the community (network) levels).
Because MaxEnt is a method of finding least-biased probability
distributions given partial knowledge about a system, these properties
need to be represented probabilistically. For example, at the species
level, MaxEnt can be used to predict the distribution of trophic levels
among species, as well as the distribution of species' vulnerability
(number of predators) and generality (number of preys). By contrast, at
the interaction level, predictions can be made on the distribution of
interaction strengths in weighted food webs. At the community level, it
can generate probability distributions of many measures of their
emerging structure and of networks themselves (i.e.~a probability
distribution that specific network configurations are realized given the
model and constraints). Overall, the potential of this method in the
study of food webs is broad. The applicability and performance of MaxEnt
mostly depend on the ecological information available and on our
capacity to find the set of state variables that best represent natural
systems and to translate them into appropriate statistical constraints.
Having a validated maximum entropy model for the system at hand allows
us to make least-biased predictions using a minimal amount of data, as
well as identify the most important ecological processes shaping that
system.

@@

We tested our analytical and heuristic MaxEnt models against open
food-web data queried from three different sources and integrated into
what we call our \emph{complete dataset}. These sources include (1)
terrestrial and aquatic food webs sampled globally and archived on the
ecological interactions database Mangal; (2) freshwater stream food webs
from New Zealand; and (3) aquatic food webs from Tuesday Lake (Michigan,
United States). First, all food webs archived on \texttt{mangal.io}
(Timothee Poisot et al. 2016; Banville, Vissault, and Poisot 2021) were
directly queried from the database (\(n = 235\)). Most ecological
networks archived on Mangal are multilayer networks, i.e.~networks that
describe different types of interactions. We kept all networks whose
interactions were mainly of predation and herbivory types, and removed
the largest network (\(S = 714\)) for computational efficiency reasons.
Then, to this set we added food webs from two different sources: the
New-Zealand dataset (\(n = 21\); Pomeranz et al. 2018) and the Tuesday
Lake dataset (\(n = 2\); Cohen, Jonsson, and Carpenter 2003). Our
complete dataset thus contained a total of \(257\) food webs. Of the
New-Zealand and Tuesday Lake datasets, \(19\) networks had data on
species' relative abundances that were used in the neutral model (Box
2). These \(19\) food webs constitute our \emph{abundance dataset},
which is a subset of our complete dataset.

>>>>

Understanding the ecological constraints that shape species interactions
networks and predicting their emerging structure are thus two
complementary aims of network ecology. This distinction between
understanding and predicting is essential when using statistical and
mathematical models in network ecology and interpreting them. On one
hand, null models help us identify potential ecological mechanisms that
drive species interactions and constrain ecological networks. Null
models generate a distribution for a target measure using a set of rules
that exclude the mechanism of interest (Fortuna and Bascompte 2006;
Delmas et al. 2019). The difference between the model and empirical data
helps us evaluate the effect of this ecological process in nature. On
the other hand, predictive models can help fill many gaps on species
interactions data. A variety of such models have recently been developed
using machine learning and other statistical tools, most of which are
presented in Strydom et al. (2021). However, given the constrained
nature of ecological networks, it is surprising that the principle of
maximum entropy, a rigorous mathematical method designed for both the
analysis and prediction of constrained systems, has been barely used in
network ecology.

@@

The principle of maximum entropy has been used in a wide range of
disciplines, from thermodynamics, chemistry and biology (Martyushev and
Seleznev 2006) to graph and network theory (e.g., Park and Newman 2004;
van der Hoorn, Lippner, and Krioukov 2018). It has also been proven
useful in ecology, e.g.~in species distribution models (Phillips,
Anderson, and Schapire 2006) and macroecological models (Harte et al.
2008; Harte and Newman 2014). In network ecology, MaxEnt has been used
to predict the degree distribution of bipartite networks from the number
of species and the number of interactions (Williams 2011) and to predict
interaction strengths between species pairs using their relative
abundances within an optimal transportation theory regularized with
information entropy (Stock, Poisot, and De Baets 2021). However, to the
best of our knowledge, MaxEnt has never been used to predict food-web
structure directly, even though food webs are among the most documented
and widespread ecological networks (Ings et al. 2009).

Food-web properties that can be derived using MaxEnt are varied and
pertain to different elements of the network (i.e.~at the species
(node), the interaction (edge) or the community (network) levels).
Because MaxEnt is a method of finding least-biased probability
distributions given partial knowledge about a system, these properties
need to be represented probabilistically. For example, at the species
level, MaxEnt can be used to predict the distribution of trophic levels
among species, as well as the distribution of species' vulnerability
(number of predators) and generality (number of preys). By contrast, at
the interaction level, predictions can be made on the distribution of
interaction strengths in weighted food webs. At the community level, it
can generate probability distributions of many measures of their
emerging structure and of networks themselves (i.e.~a probability
distribution that specific network configurations are realized given the
model and constraints). Overall, the potential of this method in the
study of food webs is broad. The applicability and performance of MaxEnt
mostly depend on the ecological information available and on our
capacity to find the right set of state variables that best represent
natural systems and to translate them into appropriate statistical
constraints. Having a validated maximum entropy model for the system at
hand allows us to make least-biased predictions using a minimal amount
of data, as well as identify the most important ecological processes
shaping that system. In other words, MaxEnt is a valuable and
underexploited tool in network ecology that allows us to better
understand and predict the structure of ecological networks.

@@

We tested our analytical and heuristic MaxEnt models against open
food-web data queried from three different sources and integrated into
what we call our \emph{complete dataset}. These sources include (1)
terrestrial and aquatic food webs sampled globally and archived on the
ecological interactions database Mangal; (2) freshwater stream food webs
from New Zealand; and (3) aquatic food webs from Tuesday Lake (Michigan,
United States). First, all food webs archived on \texttt{mangal.io}
(Timothee Poisot et al. 2016; Banville, Vissault, and Poisot 2021) were
directly queried from the database (\(N = 235\)). Most ecological
networks archived on Mangal are multilayer networks, i.e.~networks that
describe different types of interactions. We kept all networks whose
interactions were mainly of predation and herbivory types, and removed
the largest network (\(S = 714\)) for computational efficiency reasons.
Then, to this set we added food webs from two different sources: the
New-Zealand dataset (\(N = 21\); Pomeranz et al. 2018) and the Tuesday
Lake dataset (\(N = 2\); Cohen, Jonsson, and Carpenter 2003). Our
complete dataset thus contained a total of \(257\) food webs. Of the
New-Zealand and Tuesday Lake datasets, \(19\) networks had data on
species' relative abundances that were used in the neutral model (Box
2). These \(19\) food webs constitute our \emph{abundance dataset},
which is a subset of our complete dataset.
